---
title: "Transition Matrix"
output: html_document
date: "2026-04-17"
---


```{r}
library(terra)
library(gt)
library(dplyr)
library(stringr)
library(expm)
```

```{r}
# Load Data
r <- rast("SenegalData.tif")

# Add landCover for 2024
lc_2024 <- rast("C:/Users/erikg/Downloads/desert_project/LandCover_Senegal/LC.h16.2024.tif")
names(lc_2024) <- "LandCover_2024"
r <- c(r, lc_2024)
```

Splitting Markov Matrix into before and after GGW
```{r}
# Get LandCover data only.
lc_names <- names(r)[grepl("LandCover_", names(r))]
r_lc <- r[[lc_names]]

# Classes
all_classes <- unique(r_lc) %>% 
  as.vector() %>% 
  unlist() %>% 
  unique()
all_classes <- sort(all_classes[!is.na(all_classes)])

# Function subsetting data by year
get_pooled_markov <- function(rast_stack, start_year, end_year, classes) {
  n <- length(classes)
  total_counts <- matrix(0, nrow = n, ncol = n, 
                         dimnames = list(classes, classes))
  
  years <- start_year:(end_year - 1)
  
  for (y in years) {
    layer_a <- paste0("LandCover_", y)
    layer_b <- paste0("LandCover_", y + 1)
    
    if(layer_a %in% names(rast_stack) & layer_b %in% names(rast_stack)) {
      ct <- crosstab(rast_stack[[c(layer_a, layer_b)]], long = FALSE)
      
      rows <- rownames(ct)
      cols <- colnames(ct)
      total_counts[rows, cols] <- total_counts[rows, cols] + as.matrix(ct)
    }
  }
  
  row_sums <- rowSums(total_counts)
  matrix_prob <- total_counts / ifelse(row_sums == 0, 1, row_sums)
  return(matrix_prob)
}

m1 <- get_pooled_markov(r_lc, 2001, 2007, all_classes)
m2 <- get_pooled_markov(r_lc, 2008, 2024, all_classes)

# Display
render_pretty_matrix <- function(mat, title) {
  as.data.frame(mat) %>%
    mutate(From = rownames(.)) %>%
    relocate(From) %>%
    gt() %>%
    tab_header(title = md(paste0("**", title, "**"))) %>%
    fmt_number(columns = -From, decimals = 4) %>%
    data_color(
      columns = -From,
      palette = "viridis", # Professional scientific palette
      alpha = 0.8
    ) %>%
    tab_options(table.font.size = "small")
}

# Generate Results
render_pretty_matrix(m1, "Annual Transitions (2001-2007)")
render_pretty_matrix(m2, "Annual Transitions (2008-2024)")
```

Getting annual transitions
```{r}
library(terra)
library(dplyr)
library(tidyr)
library(ggplot2)

years_start <- 2001:2023

annual_transitions_long <- lapply(years_start, function(y) {
  
  layer_a <- paste0("LandCover_", y)
  layer_b <- paste0("LandCover_", y + 1)
  
  ct <- crosstab(r_lc[[c(layer_a, layer_b)]], long = TRUE)
  
  colnames(ct) <- c("From", "To", "Count")
  
  ct <- ct %>%
    group_by(From) %>%
    mutate(Prob = Count / sum(Count, na.rm = TRUE)) %>%
    ungroup() %>%
    mutate(Year = y)
  
  return(ct)
}) %>% bind_rows()

annual_transitions_long <- annual_transitions_long %>%
  mutate(Prob = replace_na(Prob, 0))
```

Edited to only two states: desert and green
Green includes all types of forest, shrubland, savanna, grassland, wetland, cropland and natural mosaic.
```{r}
# Desert mapped from 16 to 1
# Green mapped from (1-12 and 14) to 2
rcl_matrix <- matrix(c(
  15, 16, 1,
  1, 13, 2,
  13,  14, 2
), ncol = 3, byrow = TRUE)

# Reclassifying
r_binary <- classify(r_lc[[1]], rcl_matrix, others = NA)


freq(r_binary)
freq(r_lc[[2]])


names(r_binary) <- names(r_lc)


# Transitions
years_start <- 2001:2023 

total_counts <- matrix(0, nrow = 2, ncol = 2, 
                       dimnames = list(c("1", "2"), c("1", "2")))

for (y in years_start) {
  layer_a <- paste0("LandCover_", y)
  layer_b <- paste0("LandCover_", y + 1)
  
  ct <- crosstab(r_binary[[c(layer_a, layer_b)]], long = FALSE)
  
  ct_mat <- as.matrix(ct)
  
  existing_rows <- intersect(rownames(ct_mat), rownames(total_counts))
  existing_cols <- intersect(colnames(ct_mat), colnames(total_counts))
  
  if(length(existing_rows) > 0 && length(existing_cols) > 0) {
    total_counts[existing_rows, existing_cols] <- total_counts[existing_rows, existing_cols] + ct_mat[existing_rows, existing_cols]
  }
}


markov_final <- prop.table(total_counts, margin = 1)

rownames(markov_final) <- c("Barren", "Green")
colnames(markov_final) <- c("Barren", "Green")





as.data.frame(markov_final) %>%
  mutate(From = rownames(.)) %>%
  relocate(From) %>%
  gt() %>%
  tab_header(
    title = md("**Senegal: Barren vs. Green Transitions**"),
    subtitle = "Annualized Probabilities (Pooled 2001-2024)"
  ) %>%
  fmt_number(columns = -From, decimals = 5) %>%
  cols_label(From = "Starting State") %>%
  data_color(
    columns = -From,
    palette = c("#f4a582", "#92c5de")
  )
```
Senegal: Barren vs. Green Transitions
Annualized Probabilities (Pooled 2001-2024)
Starting State	Barren	Green
Barren	0.98781	0.01219
Green	0.00404	0.99596


Splitting into before and after GGW
```{r}
pre_ggw <- 2001:2006
post_ggw  <- 2008:2023

calculate_markov <- function(years, stack) {
  total_counts <- matrix(0, nrow = 2, ncol = 2, dimnames = list(c("1", "2"), c("1", "2")))
  
  for (y in years) {
    layer_t1 <- paste0("LandCover_", y)
    layer_t2 <- paste0("LandCover_", y + 1)
    
    ct <- crosstab(stack[[c(layer_t1, layer_t2)]], long = FALSE)
    ct_mat <- as.matrix(ct)
    
    rows <- intersect(rownames(ct_mat), rownames(total_counts))
    cols <- intersect(colnames(ct_mat), colnames(total_counts))
    
    if(length(rows) > 0 && length(cols) > 0) {
      total_counts[rows, cols] <- total_counts[rows, cols] + ct_mat[rows, cols]
    }
  }
  
  m_prob <- prop.table(total_counts, margin = 1)
  
  rownames(m_prob) <- c("Desert", "Vegetated")
  colnames(m_prob) <- c("Desert", "Vegetated")
  
  return(m_prob)
}

# 3. Execute the calculations
matrix_01_07 <- calculate_markov(pre_ggw, r_binary)
matrix_08_24 <- calculate_markov(post_ggw, r_binary)

# 4. Display the results
print("--- Transition Matrix: 2001 to 2007 ---")
print(round(matrix_01_07, 5))

print("--- Transition Matrix: 2008 to 2024 ---")
print(round(matrix_08_24, 5))
```
CHAD
[1] "--- Transition Matrix: 2001 to 2007 ---"
           Desert Vegetated
Desert    0.99587   0.00413
Vegetated 0.00404   0.99596
[1] "--- Transition Matrix: 2008 to 2024 ---"
           Desert Vegetated
Desert    0.99768   0.00232
Vegetated 0.00259   0.99741

SENEGAL
[1] "--- Transition Matrix: 2001 to 2007 ---"
           Desert Vegetated
Desert    0.98108   0.01892
Vegetated 0.00377   0.99623
[1] "--- Transition Matrix: 2008 to 2024 ---"
           Desert Vegetated
Desert    0.98982   0.01018
Vegetated 0.00407   0.99593



Now working with data split into populated and unpopulated.
Unpopulated defined by having less than 1 person per sq k.
```{r}

r_all <- rast("C:/Users/erikg/Downloads/desert_project/SenegalData.tif")

pop_layers <- r_all[[str_detect(names(r_all), "Pop")]]
lc_layers  <- r_all[[str_detect(names(r_all), "LandCover")]]

# Desert mapped from 16 to 1
# Green mapped from (1-12 and 14) to 2
rcl_matrix <- matrix(c(
  15, 16, 1,
  1,  13, 2,
  13, 14, 2
), ncol = 3, byrow = TRUE)

lc_binary <- classify(lc_layers, rcl_matrix, others = NA)


counts_pop   <- matrix(0, 2, 2, dimnames = list(c("1", "2"), c("1", "2")))
counts_unpop <- matrix(0, 2, 2, dimnames = list(c("1", "2"), c("1", "2")))

# Going from 2001-2020 since no data for pop past 2020
years <- 2001:2019

for (y in years) {
  cat("Analyzing transition:", y, "to", y+1, "\n")
  
  t1_name <- names(lc_binary)[str_detect(names(lc_binary), as.character(y))]
  t2_name <- names(lc_binary)[str_detect(names(lc_binary), as.character(y+1))]
  pop_name <- names(pop_layers)[str_detect(names(pop_layers), as.character(y))]
  
  is_populated <- classify(pop_layers[[pop_name]], matrix(c(-Inf, 1, 0, 1, Inf, 1), ncol=3, byrow=TRUE))
  
  pair_pop   <- mask(lc_binary[[c(t1_name, t2_name)]], is_populated, maskvalues = 0)
  pair_unpop <- mask(lc_binary[[c(t1_name, t2_name)]], is_populated, maskvalues = 1)
  
  ct_p <- as.matrix(crosstab(pair_pop, long = FALSE))
  ct_u <- as.matrix(crosstab(pair_unpop, long = FALSE))
  
  p_rows <- intersect(rownames(ct_p), rownames(counts_pop))
  p_cols <- intersect(colnames(ct_p), colnames(counts_pop))
  if(length(p_rows) > 0) {
    counts_pop[p_rows, p_cols] <- counts_pop[p_rows, p_cols] + ct_p[p_rows, p_cols]
  }
  
  u_rows <- intersect(rownames(ct_u), rownames(counts_unpop))
  u_cols <- intersect(colnames(ct_u), colnames(counts_unpop))
  if(length(u_rows) > 0) {
    counts_unpop[u_rows, u_cols] <- counts_unpop[u_rows, u_cols] + ct_u[u_rows, u_cols]
  }
}

# Final Markov tables
matrix_pop   <- prop.table(counts_pop, margin = 1)
matrix_unpop <- prop.table(counts_unpop, margin = 1)

final_names <- list(c("Desert", "Vegetated"), c("Desert", "Vegetated"))
dimnames(matrix_pop) <- final_names
dimnames(matrix_unpop) <- final_names

print("--- POPULATED MATRIX ---")
print(matrix_pop)
print("--- UNPOPULATED MATRIX (<= 1) ---")
print(matrix_unpop)
```
SENEGAL
[1] "--- POPULATED MATRIX ---"
               Desert  Vegetated
Desert    0.962225541 0.03777446
Vegetated 0.003405999 0.99659400
[1] "--- UNPOPULATED MATRIX (<= 1) ---"
              Desert   Vegetated
Desert    0.99569961 0.004300393
Vegetated 0.09252181 0.907478192

CHAD
[1] "--- POPULATED MATRIX ---"
               Desert  Vegetated
Desert    0.981899772 0.01810023
Vegetated 0.002858369 0.99714163
[1] "--- UNPOPULATED MATRIX (<= 1) ---"
              Desert    Vegetated
Desert    0.99961719 0.0003828119
Vegetated 0.06441232 0.9355876754




The code is lost, but this is the result of splitting by both time period and population.

Senegal
--- PERIOD 1: 2001-2007 ---
[1] "POPULATED:"
           Desert Vegetated
Desert    0.94517   0.05483
Vegetated 0.00314   0.99686
[1] "UNPOPULATED:"
           Desert Vegetated
Desert    0.99469   0.00531
Vegetated 0.07048   0.92952

--- PERIOD 2: 2008-2020 ---
[1] "POPULATED:"
           Desert Vegetated
Desert    0.97287   0.02713
Vegetated 0.00356   0.99644
[1] "UNPOPULATED:"
           Desert Vegetated
Desert    0.99631   0.00369
Vegetated 0.10368   0.89632


CHAD
--- PERIOD 1: 2001-2007 ---
[1] "POPULATED:"
           Desert Vegetated
Desert    0.96258   0.03742
Vegetated 0.00361   0.99639
[1] "UNPOPULATED:"
           Desert Vegetated
Desert    0.99948   0.00052
Vegetated 0.04903   0.95097

--- PERIOD 2: 2008-2020 ---
[1] "POPULATED:"
           Desert Vegetated
Desert    0.98770   0.01230
Vegetated 0.00243   0.99757
[1] "UNPOPULATED:"
           Desert Vegetated
Desert    0.99971   0.00029
Vegetated 0.09521   0.90479


40 year projection
```{r}
current_country <- "Senegal" # Change to "Chad" or "Senegal" as needed

results_senegal <- list(
  Populated = matrix(
    c(0.962225541, 0.03777446, 0.003405999, 0.99659400), 
    nrow = 2, byrow = TRUE,
    dimnames = list(c("Desert", "Vegetated"), c("Desert", "Vegetated"))
  ),
  Unpopulated = matrix(
    c(0.99569961, 0.004300393, 0.09252181, 0.907478192), 
    nrow = 2, byrow = TRUE,
    dimnames = list(c("Desert", "Vegetated"), c("Desert", "Vegetated"))
  )
)

results_chad <- list(
  Populated = matrix(
    c(0.981899772, 0.01810023, 0.002858369, 0.99714163), 
    nrow = 2, byrow = TRUE,
    dimnames = list(c("Desert", "Vegetated"), c("Desert", "Vegetated"))
  ),
  Unpopulated = matrix(
    c(0.99961719, 0.0003828119, 0.06441232, 0.9355876754), 
    nrow = 2, byrow = TRUE,
    dimnames = list(c("Desert", "Vegetated"), c("Desert", "Vegetated"))
  )
)

# picking the right country's results
results_late <- if (current_country == "Senegal") results_senegal else results_chad


r_all <- rast(paste0("C:/Users/erikg/Downloads/desert_project/", current_country, "Data.tif"))

pop_layers <- r_all[[str_detect(names(r_all), "Pop")]]
lc_layers  <- r_all[[str_detect(names(r_all), "LandCover")]]

rcl_matrix <- matrix(c(15, 16, 1, 1, 13, 2, 13, 14, 2), ncol = 3, byrow = TRUE)
lc_binary  <- classify(lc_layers, rcl_matrix, others = NA)

# get first layer
lc_2020_name <- names(lc_binary)[str_detect(names(lc_binary), "2020")]
pop_2020_name <- names(pop_layers)[str_detect(names(pop_layers), "2020")]

is_populated_2020 <- classify(pop_layers[[pop_2020_name]], matrix(c(-Inf, 1, 0, 1, Inf, 1), ncol=3, byrow=TRUE))

# extract Counts for each category
lc_pop_2020 <- mask(lc_binary[[lc_2020_name]], is_populated_2020, maskvalues = 0)
f_pop <- freq(lc_pop_2020)
counts_pop_2020 <- c(Desert = f_pop$count[f_pop$value == 1], Vegetated = f_pop$count[f_pop$value == 2])

lc_unpop_2020 <- mask(lc_binary[[lc_2020_name]], is_populated_2020, maskvalues = 1)
f_unpop <- freq(lc_unpop_2020)
counts_unpop_2020 <- c(Desert = f_unpop$count[f_unpop$value == 1], Vegetated = f_unpop$count[f_unpop$value == 2])


project_40_years <- function(start_counts, transition_matrix) {
  years <- 0:40
  results <- data.frame(Year = 2020 + years, Desert = 0, Vegetated = 0)
  
  v <- matrix(start_counts, nrow = 1)
  for (i in seq_along(years)) {

    future_v <- v %*% (transition_matrix %^% (i - 1))
    results[i, c("Desert", "Vegetated")] <- as.numeric(future_v)
  }
  return(results)
}

# run projections
proj_human <- project_40_years(counts_pop_2020, results_late$Populated)
proj_wild  <- project_40_years(counts_unpop_2020, results_late$Unpopulated)

# plotting
proj_human$Zone <- "Populated (Human)"
proj_wild$Zone  <- "Unpopulated (Wild)"
full_projection <- rbind(proj_human, proj_wild)


plot_df <- full_projection %>%
  pivot_longer(cols = c(Desert, Vegetated), names_to = "Class", values_to = "Pixels")

ggplot(plot_df, aes(x = Year, y = Pixels, color = Class, linetype = Zone)) +
  geom_line(linewidth = 1) +
  scale_color_manual(values = c("Desert" = "#d8b365", "Vegetated" = "#1a9850")) +
  theme_minimal() +
  labs(title = paste0(current_country, ": 2020-2060 Projection"),
       subtitle = "Human-managed areas vs. Wilderness dynamics",
       y = "Total Pixel Count",
       x = "Year") +
  facet_wrap(~Zone, scales = "free_y")
```
